01 - TreehopperFitness - Quality control and data prep

### 2023-08-25 scripts/Publilia-01-qc.Rmd
minor edits to ensure scripts 01-03 run after overhaul of strats_data

---
title: '01 - TreehopperFitness: Quality control'
author: "Micah Fletcher"
date: '`r Sys.Date()`'
output: 
  rmdformats::robobook:
  fig_width: 9
toc_depth: 3
---

@@

## 7.1 - Adding second gen adult counts per leaf to clutch status stats
```{r}
# counting adults
eA_status_v1 <- All_data_v6 %>%
  group_by(row_number()) %>% # sum of newly exclosed adults
  filter(any(!is.na(c(New_eA, Old_eA)))) %>%
  mutate(n_newA = ifelse(is.na(New_eA),
                         NA,
                         sum(as.numeric(regmatches(New_eA, gregexpr("[0-9]+", New_eA))[[1]]))),
         n_oldA = ifelse(is.na(Old_eA),
                         NA,
                         sum(as.numeric(regmatches(Old_eA, gregexpr("[0-9]+", Old_eA))[[1]]))),
         n_A = sum(n_newA, n_oldA, na.rm = TRUE)) %>%
  ungroup() %>%
  select(Date, location, n_newA, n_oldA, n_A)

@@

## 7.2 - Manually check nymph origins for revision
First create a table of all entried from clutches that were not flagged in the field as having ambiguous origin (e.g. "leaf 6 or 9") in cases were more than one clutch was on the plant. These were revised manually to specify a leaf of origin for the nymphs, where possible.
```{r}
NOr_status_resolveable_all_unambig_clutches <- All_data_v6_clutch_status_stats_v2 %>% 
  select(Date, Dorm, Plant, Leaf, location, 
         egg_num, N1, N2, N5, n_A, nymph_origin, initiation_date, hatch_date,
         any_NOr_ambig_loc, Omit_for_abandon_time, never_hatched,
         n_clutches_on_plant) %>%
  # keep only entries with some nymphs or adults on a leaf
  # OR entries for leaves where clutches have been initiated
  group_by(row_number()) %>%
  filter(!is.na(Leaf) & any(N1 > 0, N2 > 0, N5 > 0, n_A > 0) | 
           Date >= initiation_date) %>% # nrow = 17944
  # flag entries where the nymph origin is text (typically ambiguous, e.g. "8 or 9")
  mutate(NOr_ambig_entry = !is.na(nymph_origin) & is.na(as.numeric(nymph_origin))) %>% 
  ungroup() %>%
  # get rid of duplicate rows from multiple females in the same location
  distinct(Date, location, .keep_all = TRUE) %>% # nrow = 15674
  group_by(Dorm, Plant) %>%
    # filter out any plants with a clutch that was omitted for abandon time
    filter(!any(!is.na(Omit_for_abandon_time))) %>%
    # flag plants with any ambiguous entries (e.g. "8 or 9") skip for checking
    mutate(NOr_ambig_plant = any(NOr_ambig_entry)) %>%
  ungroup()

@@

## 7.3 - Implement revisions by creating the new variables NOr_revised and NOr_rev_loc
```{r}
# if nymph origin was clearly resolveable, the revisions were made in NOr_manual_revisions
# updating NOr_status with revisions so that it can be left_joined to other data and
# counts of nymphs and adults on NOr can be calculated
NOr_status_revised_all_unambig_clutches <- NOr_status_resolveable_all_unambig_clutches %>%
  left_join(NOr_manual_revisions, by = c("Date", "location")) %>%
  group_by(Dorm, Plant) %>%
    mutate(lone_hatcher_leaf = ifelse(length(unique(Leaf[!is.na(initiation_date)])) == 1,
                                 unique(Leaf[!is.na(initiation_date)]),
                                 NA),
           NOr_revised = ifelse(!is.na(NOr_revised_manual),
                                       NOr_revised_manual,
                                ifelse(any(!is.na(lone_hatcher_leaf)),
                                       lone_hatcher_leaf,
                                       # if no clutches hatched on the plant, NOr_revised = Leaf
                                       # for leaves with eggs
                                       ifelse(!any(!is.na(hatch_date)) & !is.na(egg_num),
                                              Leaf,
                                              NA)))) %>%
  ungroup() %>%  
  group_by(location) %>%
    mutate(NOr_revised = ifelse(any(NOr_revised == Leaf) & is.na(NOr_revised),
                                Leaf,#unique(NOr_revised[!is.na(NOr_revised)]),
                                NOr_revised)) %>%
  ungroup() %>%
  mutate(NOr_rev_loc = ifelse(is.na(NOr_revised),
                              NA,
                              paste0(Dorm, Plant, "_", NOr_revised))) %>%
  select(Date, location, NOr_revised, NOr_rev_loc) %>%
  filter(!is.na(NOr_rev_loc))

@@

```{r}
NOr_status_resolveable_all_unambig_clutches <- All_data_v6_clutch_status_stats_v2 %>% 
  select(Date, Dorm, Plant, Leaf, location, 
         egg_num, N1, N2, N5, n_A, nymph_origin, initiation_date, hatch_date,
         any_NOr_ambig_loc, Omit_for_abandon_time, never_hatched,
         n_clutches_on_plant) %>%
  # keep only entries with some nymphs or adults on a leaf
  # OR entries for leaves where clutches have been initiated
  group_by(row_number()) %>%
  filter(!is.na(Leaf) & any(N1 > 0, N2 > 0, N5 > 0, n_A > 0) | 
           Date >= initiation_date) %>% # nrow = 17944
  # flag entries where the nymph origin is text (typically ambiguous, e.g. "8 or 9")
  mutate(NOr_ambig_entry = !is.na(nymph_origin) & is.na(as.numeric(nymph_origin))) %>% 
  ungroup() %>%
  # get rid of duplicate rows from multiple females in the same location
  distinct(Date, location, .keep_all = TRUE) %>% # nrow = 15674
  group_by(Dorm, Plant) %>%
    # filter out any plants with a clutch that was omitted for abandon time
    filter(!any(!is.na(Omit_for_abandon_time))) %>%
    # flag plants with any ambiguous entries (e.g. "8 or 9") skip for checking
    mutate(NOr_ambig_plant = any(NOr_ambig_entry)) %>%
  ungroup()

@@

- Filtered out clutches with outlier numbers of N1 or N5 etc at early or late clutch ages, which were assumed to be cases of immigration that weren't caught during manual nymph origin filtering
```{r}
# Filter to only include curated list of clutches with unambiguous nymph counts after accounting for immigration to non-natal leaves and then sum all nymphs (including emigrants) for each clutch
# I also added clutch_age variables to help identify clutches that were started too close to the end of sampling to be useful
clutch_unambig_NOr_revised_ENAa_summary_v2 <- All_data_v6_clutch_status_stats_v2 %>%
  right_join(NOr_status_revised_all_unambig_clutches_v2, by = c("Date", "location")) %>%
  select(Date, NOr_rev_loc, location, egg_num, N1, N2, N5, n_A, ants_L) %>%
    # remove any duplicates - I need to figure out why there are dups later but for 4th yr talk
  distinct(Date, NOr_rev_loc, location, egg_num, N1, N2, N5, n_A, ants_L, .keep_all = TRUE) %>%
  mutate(egg_num_NOr = ifelse(NOr_rev_loc == location, egg_num, NA)) %>%
  group_by(Date, NOr_rev_loc) %>%
    summarize(eggs_NOr = first(egg_num_NOr[!is.na(egg_num_NOr)]), # only use eggs from origin leaf!
              N1_NOr = sum(N1, na.rm= TRUE),
              N2_NOr = sum(N2, na.rm= TRUE),
              N5_NOr = sum(N5, na.rm= TRUE),
              N_NOr = sum(N1, N2, N5, na.rm = TRUE),
              A_NOr = sum(n_A, na.rm= TRUE),
              N5toA_NOr = sum(N5, n_A, na.rm= TRUE),
              ants_NOr = sum(ants_L, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(filter(clutch_status_guarding_summary_v2, mom_at_init), by = c("NOr_rev_loc" = "location")) %>%
  mutate(clutch_age1 = Date - initiation_date,
         clutch_age2 = ceiling((Date - initiation_date)/2)*2,
         clutch_age4 = ceiling((Date - initiation_date)/4)*4,
         after_hatch1 = Date - hatch_date,
         after_hatch2 = ceiling((Date - hatch_date)/2)*2,
         after_hatch4 = ceiling((Date - hatch_date)/4)*4) %>%
  select(Date, NOr_rev_loc, clutch_age1, after_hatch1, eggs_NOr, 
         N1_NOr, N2_NOr, N5_NOr, N_NOr, A_NOr, N5toA_NOr, ants_NOr,
         init_fem = Female_ID,
         initiation_date, hatch_date, guard_duration, relative_desert_date,
         clutch_initiated_order, females_removed, ants_removed, 
         clutch_age2, clutch_age4, after_hatch2, after_hatch4) #%>%
  # filter(initiation_date <= as_date("2019-06-10"),
  #        clutch_age4 <= 68)

@@

## 10.2 - Calculate median of top 5 egg counts in lieu of max egg counts
Taking the median (or mean) of the top 5 egg_num counts yields an estimate of max egg num that is more robust to outliers/miscounts
```{r}
tmp_t5eggs_data <- clutch_status_init_hatch_labels_v1 %>%
  mutate(est_hatch = as_date(ifelse(is.na(hatch_date),
                                 initiation_date + 20,
                                 hatch_date)),
         clutch_age = as.numeric(Date-initiation_date)) %>%
  # remove duplicate egg counts due to multiple females at a location on a day
  distinct(location, Date, initiation_date, est_hatch, clutch_age,
           egg_num, egg_len) %>%
  group_by(location) %>%
    arrange(desc(egg_num)) %>%
    mutate(top5egg_counts = rank(desc(egg_num), ties.method = "first") <= 5,
           top5egg_lens = rank(desc(egg_len), ties.method = "first") <= 5) %>%
    #filter(location == "A13_11") %>%
    select(location, Date, initiation_date, est_hatch, clutch_age,
           egg_num, egg_len, top5egg_counts, top5egg_lens) %>%
    mutate(med_t5eggs = median(ifelse(top5egg_counts,egg_num,NA), na.rm=T),
           mean_t5eggs = mean(ifelse(top5egg_counts,egg_num,NA), na.rm=T),
           max_n_eggs = max(ifelse(top5egg_counts,egg_num,NA), na.rm=T),
           med_l_t5eggs = median(ifelse(top5egg_lens,egg_len,NA), na.rm=T),
           mean_l_t5eggs = mean(ifelse(top5egg_lens,egg_len,NA), na.rm=T),
           max_l_eggs = max(ifelse(top5egg_lens,egg_len,NA), na.rm=T)) %>%
  ungroup()

>>>>

---
title: '01 - TreehopperFitness - Quality control and data prep'
author: "Micah Fletcher"
date: '`r Sys.Date()`'
output: 
  rmdformats::robobook:
  fig_width: 9
toc_depth: 3
---

@@

## 7.1 - Adding second gen adult counts per leaf to clutch status stats
```{r, warning=FALSE}
# counting adults
eA_status_v1 <- All_data_v6 %>%
  group_by(row_number()) %>% # sum of newly exclosed adults
  filter(any(!is.na(c(New_eA, Old_eA)))) %>%
  mutate(n_newA = ifelse(is.na(New_eA),
                         NA,
                         sum(as.numeric(regmatches(New_eA, gregexpr("[0-9]+", New_eA))[[1]]))),
         n_oldA = ifelse(is.na(Old_eA),
                         NA,
                         sum(as.numeric(regmatches(Old_eA, gregexpr("[0-9]+", Old_eA))[[1]]))),
         n_A = sum(n_newA, n_oldA, na.rm = TRUE)) %>%
  ungroup() %>%
  select(Date, location, n_newA, n_oldA, n_A)

@@

## 7.2 - Manually check nymph origins for revision
First create a table of all entried from clutches that were not flagged in the field as having ambiguous origin (e.g. "leaf 6 or 9") in cases were more than one clutch was on the plant. These were revised manually to specify a leaf of origin for the nymphs, where possible.
```{r, warning=FALSE}
NOr_status_resolveable_all_unambig_clutches <- All_data_v6_clutch_status_stats_v2 %>% 
  select(Date, Dorm, Plant, Leaf, location, 
         egg_num, N1, N2, N5, n_A, nymph_origin, initiation_date, hatch_date,
         any_NOr_ambig_loc, Omit_for_abandon_time, never_hatched,
         n_clutches_on_plant) %>%
  # keep only entries with some nymphs or adults on a leaf
  # OR entries for leaves where clutches have been initiated
  group_by(row_number()) %>%
  filter(!is.na(Leaf) & any(N1 > 0, N2 > 0, N5 > 0, n_A > 0) | 
           Date >= initiation_date) %>% # nrow = 17944
  # flag entries where the nymph origin is text (typically ambiguous, e.g. "8 or 9")
  mutate(NOr_ambig_entry = !is.na(nymph_origin) & is.na(as.numeric(nymph_origin))) %>% 
  ungroup() %>%
  # get rid of duplicate rows from multiple females in the same location
  distinct(Date, location, .keep_all = TRUE) %>% # nrow = 15674
  group_by(Dorm, Plant) %>%
    # filter out any plants with a clutch that was omitted for abandon time
    filter(!any(!is.na(Omit_for_abandon_time))) %>%
    # flag plants with any ambiguous entries (e.g. "8 or 9") skip for checking
    mutate(NOr_ambig_plant = any(NOr_ambig_entry)) %>%
  ungroup()

@@

## 7.3 - Implement revisions by creating the new variables NOr_revised and NOr_rev_loc
```{r, warning=FALSE}
# if nymph origin was clearly resolveable, the revisions were made in NOr_manual_revisions
# updating NOr_status with revisions so that it can be left_joined to other data and
# counts of nymphs and adults on NOr can be calculated
NOr_status_revised_all_unambig_clutches <- NOr_status_resolveable_all_unambig_clutches %>%
  left_join(NOr_manual_revisions, by = c("Date", "location")) %>%
  group_by(Dorm, Plant) %>%
    mutate(lone_hatcher_leaf = ifelse(length(unique(Leaf[!is.na(initiation_date)])) == 1,
                                 unique(Leaf[!is.na(initiation_date)]),
                                 NA),
           NOr_revised = ifelse(!is.na(NOr_revised_manual),
                                       NOr_revised_manual,
                                ifelse(any(!is.na(lone_hatcher_leaf)),
                                       lone_hatcher_leaf,
                                       # if no clutches hatched on the plant, NOr_revised = Leaf
                                       # for leaves with eggs
                                       ifelse(!any(!is.na(hatch_date)) & !is.na(egg_num),
                                              Leaf,
                                              NA)))) %>%
  ungroup() %>%  
  group_by(location) %>%
    mutate(NOr_revised = ifelse(any(NOr_revised == Leaf) & is.na(NOr_revised),
                                Leaf,#unique(NOr_revised[!is.na(NOr_revised)]),
                                NOr_revised)) %>%
  ungroup() %>%
  mutate(NOr_rev_loc = ifelse(is.na(NOr_revised),
                              NA,
                              paste0(Dorm, Plant, "_", NOr_revised))) %>%
  select(Date, location, NOr_revised, NOr_rev_loc) %>%
  filter(!is.na(NOr_rev_loc))

@@

```{r, warning=FALSE}
NOr_status_resolveable_all_unambig_clutches <- All_data_v6_clutch_status_stats_v2 %>% 
  select(Date, Dorm, Plant, Leaf, location, 
         egg_num, N1, N2, N5, n_A, nymph_origin, initiation_date, hatch_date,
         any_NOr_ambig_loc, Omit_for_abandon_time, never_hatched,
         n_clutches_on_plant) %>%
  # keep only entries with some nymphs or adults on a leaf
  # OR entries for leaves where clutches have been initiated
  group_by(row_number()) %>%
  filter(!is.na(Leaf) & any(N1 > 0, N2 > 0, N5 > 0, n_A > 0) | 
           Date >= initiation_date) %>% # nrow = 17944
  # flag entries where the nymph origin is text (typically ambiguous, e.g. "8 or 9")
  mutate(NOr_ambig_entry = !is.na(nymph_origin) & is.na(as.numeric(nymph_origin))) %>% 
  ungroup() %>%
  # get rid of duplicate rows from multiple females in the same location
  distinct(Date, location, .keep_all = TRUE) %>% # nrow = 15674
  group_by(Dorm, Plant) %>%
    # filter out any plants with a clutch that was omitted for abandon time
    filter(!any(!is.na(Omit_for_abandon_time))) %>%
    # flag plants with any ambiguous entries (e.g. "8 or 9") skip for checking
    mutate(NOr_ambig_plant = any(NOr_ambig_entry)) %>%
  ungroup()

@@

- Filtered out clutches with outlier numbers of N1 or N5 etc at early or late clutch ages, which were assumed to be cases of immigration that weren't caught during manual nymph origin filtering
```{r, warning=FALSE}
# Filter to only include curated list of clutches with unambiguous nymph counts after accounting for immigration to non-natal leaves and then sum all nymphs (including emigrants) for each clutch
# I also added clutch_age variables to help identify clutches that were started too close to the end of sampling to be useful
clutch_unambig_NOr_revised_ENAa_summary_v2 <- All_data_v6_clutch_status_stats_v2 %>%
  right_join(NOr_status_revised_all_unambig_clutches_v2, by = c("Date", "location")) %>%
  select(Date, NOr_rev_loc, location, egg_num, N1, N2, N5, n_A, ants_L) %>%
    # remove any duplicates - I need to figure out why there are dups later but for 4th yr talk
  distinct(Date, NOr_rev_loc, location, egg_num, N1, N2, N5, n_A, ants_L, .keep_all = TRUE) %>%
  mutate(egg_num_NOr = ifelse(NOr_rev_loc == location, egg_num, NA)) %>%
  group_by(Date, NOr_rev_loc) %>%
    summarize(eggs_NOr = first(egg_num_NOr[!is.na(egg_num_NOr)]), # only use eggs from origin leaf!
              N1_NOr = sum(N1, na.rm= TRUE),
              N2_NOr = sum(N2, na.rm= TRUE),
              N5_NOr = sum(N5, na.rm= TRUE),
              N_NOr = sum(N1, N2, N5, na.rm = TRUE),
              A_NOr = sum(n_A, na.rm= TRUE),
              N5toA_NOr = sum(N5, n_A, na.rm= TRUE),
              ants_NOr = sum(ants_L, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(filter(clutch_status_guarding_summary_v2, mom_at_init), by = c("NOr_rev_loc" = "location")) %>%
  mutate(clutch_age1 = Date - initiation_date,
         clutch_age2 = ceiling((Date - initiation_date)/2)*2,
         clutch_age4 = ceiling((Date - initiation_date)/4)*4,
         after_hatch1 = Date - hatch_date,
         after_hatch2 = ceiling((Date - hatch_date)/2)*2,
         after_hatch4 = ceiling((Date - hatch_date)/4)*4) %>%
  select(Date, NOr_rev_loc, clutch_age1, after_hatch1, eggs_NOr, 
         N1_NOr, N2_NOr, N5_NOr, N_NOr, A_NOr, N5toA_NOr, ants_NOr,
         init_fem = Female_ID,
         initiation_date, hatch_date, guard_duration, relative_desert_date,
         clutch_initiated_order, females_removed, ants_removed, 
         clutch_age2, clutch_age4, after_hatch2, after_hatch4) #%>%
  # filter(initiation_date <= as_date("2019-06-10"),
  #        clutch_age4 <= 68)

@@

## 10.2 - Calculate median of top 5 egg counts in lieu of max egg counts
Taking the median (or mean) of the top 5 egg_num counts yields an estimate of max egg num that is more robust to outliers/miscounts
```{r, warning=FALSE}
tmp_t5eggs_data <- clutch_status_init_hatch_labels_v1 %>%
  mutate(est_hatch = as_date(ifelse(is.na(hatch_date),
                                 initiation_date + 20,
                                 hatch_date)),
         clutch_age = as.numeric(Date-initiation_date)) %>%
  # remove duplicate egg counts due to multiple females at a location on a day
  distinct(location, Date, initiation_date, est_hatch, clutch_age,
           egg_num, egg_len) %>%
  group_by(location) %>%
    arrange(desc(egg_num)) %>%
    mutate(top5egg_counts = rank(desc(egg_num), ties.method = "first") <= 5,
           top5egg_lens = rank(desc(egg_len), ties.method = "first") <= 5) %>%
    #filter(location == "A13_11") %>%
    select(location, Date, initiation_date, est_hatch, clutch_age,
           egg_num, egg_len, top5egg_counts, top5egg_lens) %>%
    mutate(med_t5eggs = median(ifelse(top5egg_counts,egg_num,NA), na.rm=T),
           mean_t5eggs = mean(ifelse(top5egg_counts,egg_num,NA), na.rm=T),
           max_n_eggs = max(ifelse(top5egg_counts,egg_num,NA), na.rm=T),
           med_l_t5eggs = median(ifelse(top5egg_lens,egg_len,NA), na.rm=T),
           mean_l_t5eggs = mean(ifelse(top5egg_lens,egg_len,NA), na.rm=T),
           max_l_eggs = max(ifelse(top5egg_lens,egg_len,NA), na.rm=T)) %>%
  ungroup()
